The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2024-03-05 r_r/manuscript_rihe_r_r.tex

We situate this study within scholarship on nonresident enrollment at public universities. Author (XXXXa) analyzed recruiting visits to public high schools made by 15 public research universities in 2017. The authors found that 12 of the 15 universities made more visits to out-of-state schools than in-state schools. These out-of-state visits focused on affluent, predominantly white public schools. Set against the scholarship about declining state appropriations and growing nonresident enrollment (Author, XXXXb, XXXXc), these results suggest that many public research universities prioritize enrollment from wealthy out-of-state students who can afford to pay nonresident tuition prices.

A separate, multidisciplinary literature has analyzed linkages betwen private school and college access (Bryk, Lee, \& Holland, 1993; Chetty, Deming, \& Friedman, 2023; Khan, 2011; Kingston \& Lewis, 1990). This research observes strong links -- and student flows -- between private high schools and selective private universities. Private schools enroll many students who can afford private university tuition (Stevens, 2007). To date, scholarship on nonresident enrollment by public universities has progressed in isolation from scholarship on links between private schools and colleges.

@@

Although the literature on nonresident enrollment developed in isolation from the literature on private school to private university linkages, both phenomena are part of a broader story about the changing nature of school segregation and the competition for affluent, traditional college students. Drawing from resource dependence theory, as public research universities become more reliant, they have a financial incentive to target out-of-state private schools, which tend to enroll affluent students that can pay high tuition prices. Although a handful of public research universities have relied on nonresident enrollment for many decades (e.g., University of Michigan, University of Delaware), public research universities began substantially increasing nonresident enrollment en masse around 2005 (Author, XXXXb). By contrast, selective private universities have long relied on enrollment from private schools (Persell \& Cookson, 1985; Stevens, 2007). Therefore, the pursuit of out-of-state private school students by public research universities is a relatively recent encroachment on the historic territory of selective private universities and in-state flagships. Speaking about this competition, an enrollment management consultant said, ``\,`Everybody wants to go to the magic island of full pay students, but it's rapidly shrinking real estate'\,'' (Rivard, 2013). By analyzing the network of recruiting visits, we show whether public research universities and selective private universities are competing for students at similar sets of private schools. This may be direct competition for the same student or vying to be the second choice for students who are rejected by their top choice.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-RN4758}
Author. (XXXXa).

\bibitem[\citeproctext]{ref-list_biz}
Author. (XXXXe).

\bibitem[\citeproctext]{ref-list_empirics}
Author. (XXXXf).

\bibitem[\citeproctext]{ref-RN3753}
Author. (XXXXb).

\bibitem[\citeproctext]{ref-RN3685}
Author. (XXXXc).

\bibitem[\citeproctext]{ref-RN4290}
Author. (XXXXd).

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-RN4758}
Author. (XXXXa).

\bibitem[\citeproctext]{ref-list_biz}
Author. (XXXXe).

\bibitem[\citeproctext]{ref-list_empirics}
Author. (XXXXf).

\bibitem[\citeproctext]{ref-RN3753}
Author. (XXXXb).

\bibitem[\citeproctext]{ref-RN3685}
Author. (XXXXc).

\bibitem[\citeproctext]{ref-RN4290}
Author. (XXXXd).

>>>>

We situate this study within scholarship on nonresident enrollment at public universities. Salazar, Jaquette, \& Han (2021) analyzed recruiting visits to public high schools made by 15 public research universities in 2017. The authors found that 12 of the 15 universities made more visits to out-of-state schools than in-state schools. These out-of-state visits focused on affluent, predominantly white public schools. Set against the scholarship about declining state appropriations and growing nonresident enrollment (Jaquette \& Curs, 2015; Jaquette, Curs, \& Posselt, 2016), these results suggest that many public research universities prioritize enrollment from wealthy out-of-state students who can afford to pay nonresident tuition prices.

A separate, multidisciplinary literature has analyzed linkages between private school and college access (Bryk, Lee, \& Holland, 1993; Chetty, Deming, \& Friedman, 2023; Khan, 2011; Kingston \& Lewis, 1990). This research observes strong links -- and student flows -- between private high schools and selective private universities. Private schools enroll many students who can afford private university tuition (Stevens, 2007). To date, scholarship on nonresident enrollment by public universities has progressed in isolation from scholarship on links between private schools and colleges.

@@

Although the literature on nonresident enrollment developed in isolation from the literature on private school to private university linkages, both phenomena are part of a broader story about the changing nature of school segregation and the competition for affluent, traditional college students. Drawing from resource dependence theory, as public research universities become more reliant, they have a financial incentive to target out-of-state private schools, which tend to enroll affluent students that can pay high tuition prices. Although a handful of public research universities have relied on nonresident enrollment for many decades (e.g., University of Michigan, University of Delaware), public research universities began substantially increasing nonresident enrollment en masse around 2005 (Jaquette \& Curs, 2015). By contrast, selective private universities have long relied on enrollment from private schools (Persell \& Cookson, 1985; Stevens, 2007). Therefore, the pursuit of out-of-state private school students by public research universities is a relatively recent encroachment on the historic territory of selective private universities and in-state flagships. Speaking about this competition, an enrollment management consultant said, ``\,`Everybody wants to go to the magic island of full pay students, but it's rapidly shrinking real estate'\,'' (Rivard, 2013). By analyzing the network of recruiting visits, we show whether public research universities and selective private universities are competing for students at similar sets of private schools. This may be direct competition for the same student or vying to be the second choice for students who are rejected by their top choice.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-RN4953}
Borgatti, S. P. (1998). Ego networks. Retrieved from \url{http://www.analytictech.com/networks/egonet.htm}

@@

\bibitem[\citeproctext]{ref-RN4290}
Curs, B. R., \& Jaquette, O. (2017). Crowded out? The effect of nonresident enrollment on resident access to public research universities. \emph{Educational Evaluation and Policy Analysis}, \emph{39}(4), 644--669. Journal Article. \url{https://doi.org/10.3102/0162373717704719}

@@

\bibitem[\citeproctext]{ref-RN3753}
Jaquette, O., \& Curs, B. R. (2015). Creating the out-of-state university: Do public universities increase nonresident freshman enrollment in response to declining state appropriations? \emph{Research in Higher Education}, \emph{56}(6), 535--565. \url{https://doi.org/10.1007/s11162-015-9362-2}

\bibitem[\citeproctext]{ref-RN3685}
Jaquette, O., Curs, B. R., \& Posselt, J. R. (2016). Tuition rich, mission poor: Nonresident enrollment growth and the. \emph{Journal of Higher Education}, \emph{87}(5), 635--673. \url{https://doi.org/10.1353/jhe.2016.0025}

\bibitem[\citeproctext]{ref-list_biz}
Jaquette, O., Salazar, K. G., \& Martin, P. (2022). \emph{The student list business: Primer and market dynamics}. The Institute for College Access; Success. Retrieved from \url{https://ticas.org/wp-content/uploads/2022/09/The-Student-List-Business_-Primer-and-Market-Dynamics.pdf}

@@

\bibitem[\citeproctext]{ref-RN4758}
Salazar, K. G., Jaquette, O., \& Han, C. (2021). Coming soon to a neighborhood near you? Off-campus recruiting by public research universities. \emph{American Educational Research Journal}, \emph{58}(6), 1270--1314. \url{https://doi.org/10.3102/00028312211001810}

\bibitem[\citeproctext]{ref-list_empirics}
Salazar, K. G., Jaquette, O., \& Han, C. (2022). \emph{Geodemographics of student list purchases: A first look}. The Institute for College Access; Success. Retrieved from \url{https://ticas.org/wp-content/uploads/2022/09/Geodemographics-of-Student-List-Purchases_A-First-Look.pdf}

@@
